---
title: "Georgia Recruit Analysis"
author: "Jeremy Harris"
date: "May 9, 2020"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Georgia Football Recruiting Analysis (Past 10 Years)
I'm going to take a look at a few excel sheets that have all of the high school commits to college from the state of Georgia for the past 10 years. The goal is to provide the best potential pipeline for WVU to establish that will yield the highest caliber of athlete that will potentially commit to WVU. 

```{r, include = FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(readxl)
library(caret)
library(OneR)
library(cluster)
library(mltools)
library(jsonlite)
library(gridExtra)
library(maps)
library(ggmap)
library(mapdata)

#load up the data from the three different tabs in the provided excel file
orig_data <- read_excel("GA_recruiting_data.xlsx", sheet = 1, na = "0")
#school_class <- read_excel("GA_recruiting_data.xlsx", sheet = 2)
#all_conf_data <- read_excel("GA_recruiting_data.xlsx", sheet = 3)
```

```{r, include = FALSE}
#tidy up the column names
recruiting_data <- orig_data

#change all-conf/nfl to factor with 1 an 0
recruiting_data$`All Conf/NFL` <- sub("X", "1", recruiting_data$`All Conf/NFL`)
recruiting_data$`All Conf/NFL`[is.na(recruiting_data$`All Conf/NFL`)] <- 0

#rename class & year - they had the same name
names(recruiting_data)[1] <- "year"
names(recruiting_data)[6] <- "hs.class"

#make tidy column names and rename columns
c_names <- colnames(recruiting_data) %>%
  make.names() 
colnames(recruiting_data) <- c_names

#remove na from dataset on 247 natl rankings
recruiting_data$X247.Nat. <- sub("NA", "0", recruiting_data$X247.Nat.) %>%
  as.numeric()

#create factors for appropriate variables
c_names2 <- colnames(recruiting_data[c(1,4:9,12,17)])
recruiting_data[c_names2] <- lapply(recruiting_data[c_names2], factor)

#remove letter form hs.class
recruiting_data$hs.class <- gsub('.{1}$', '', recruiting_data$hs.class) %>%
  as.factor()

#add in state of college and in state / out of state factor
recruiting_data$coll.state <- as.factor(substr(recruiting_data$Signed, 1, 2))

recruiting_data <- recruiting_data %>%
  mutate(in.state = ifelse((coll.state == "GA"), 1, 0),
         in.sec = ifelse((Conf. == "SEC"), 1, 0))
```

##Look At Basic Plots for Inference
I just want to look at rough plots to get an idea of how this data shapes up before I start my analysis. 

```{r, echo = FALSE}
#Conferences of commits
plot(recruiting_data$Conf.)

#hs class of commits
plot(recruiting_data$hs.class)
```

##Create "Like" WVU Category
I want to create a category that contains schools that are "like" WVU. I'll use the data I have with a little outside research to lump schools together in terms of location, conference and then final season rankings. I'll use this new category for the analysis to determine which commits would have been more likely to attend WVU with a statistical probability. From there, I will analyze the detailed information of the commits to find commonality with things such as school size, school location and recruit 247 rating. 

```{r, include = FALSE, warning = FALSE}
#like_wvu variable will be created by choosing Big 12
#and states surrounding and including WV

#add id so I can join back data later
recruiting_data$id <- seq(1:length(recruiting_data$year))

wv_rec <- recruiting_data %>%
  filter(coll.state == "WV" | coll.state =="PA" | coll.state == "MD" |
           coll.state == "VA" | coll.state == "KY" | coll.state == "OH" |
           Conf. == "B12") %>%
  mutate(like_wvu = 1) %>%
  as.data.frame()

wv_id <- as.numeric(wv_rec$id)

#join the like_wvu column back to the recruiting data
recruit_all <- recruiting_data %>%
  mutate(like_wvu = 0) %>%
  as.data.frame()

recruit_all[wv_id,]$like_wvu <- 1

#setup new variables as factors
recruit_all$like_wvu <- as.factor(recruit_all$like_wvu)
recruit_all$in.state <- as.factor(recruit_all$in.state)
recruit_all$in.sec <- as.factor(recruit_all$in.sec)
```

I can see that there are 198 of the 1,223 total commits narrowed down just by selecting the Big 12 and the surrounding states.

##Verify Predictions and Analyze the Variables
Now I will see how well I can build a model that will predict whether a recruit is likely to go to a like_wvu school. I will analyze the findings to determine what the driving factors are as well as look to narrow down locations of interest for WVU.

```{r, include = FALSE}
#setup data for predictions by selecting necessary columns
#I have to also take out any variables that made up the column "like_wvu" or
#my model will severly overfit
pred_data <- recruit_all %>%
  select(-id, -First, -Last, -State, -Conf., -Signed, -coll.state, -in.state, -in.sec, -High.School, -All.Conf.NFL)

col_scale <- c(4,5,7,8,9,10)
# 
# test <- pred_data
# test[,col_scale] <- lapply(test[,col_scale], scale)


#setup training/testing data sets with caret
#setup cross-validation with 10 folds
fitControl1 <- trainControl(method = "cv", number = 10)

#setup test and train variables
set.seed(333)
training1 <- createDataPartition(pred_data$like_wvu, p = .75)$Resample1
train_data1 <- pred_data[training1,]
test_data1 <- pred_data[-training1, ]

#setup elasticnet model
mod1_expandEN = expand.grid(lambda=seq(0.005, .03, .001), alpha=seq(0.1, 1, 0.1))
set.seed(333)
mod1.EN <- train(like_wvu ~. + X247.Star*hs.class , data = train_data1, method = "glmnet", trControl = fitControl1, family="binomial",
                 tuneGrid = mod1_expandEN)
mod1.EN
#best results: alpha = .1 & lambda = .03 --> accuracy = 83.8% accuracy

#look at coefficient results
coef(mod1.EN$finalModel, mod1.EN$bestTune$lambda)
```

Looking at the raw numbers and verifying with a variable importance output, I can see that the following variables have the highest influence on my results: 

* **HS Size**: We're going to see better results with 4A & 6A. We'll see the worst results with 7A.
* **Positions**: We'll see the best results with QB, RB, OLB & WR. Worst results with FB & PK.
* **247 Rating**: Our best star rating is 3 and 5. Worst rating is 4 star.

##More Initial Data Analysis
I create a few dataframes here to take a look at different variables that might be important. 

```{r, echo = FALSE}
#predict on my testing set of data
preds1.EN <- predict(mod1.EN, newdata = test_data1)
confusionMatrix(data = preds1.EN, reference = test_data1$like_wvu)
```

##All-Conf / NFL Feature Importance
I want to gain more insight into the variables that lead to a player becoming an All-Conference or NFL player. There is a lot to take in here that cannot be captured in the data - however, I was able to determine certain attributes that lead to a more likely scenario for success. 
```{r, include = FALSE}
# pred_data2 <- recruit_all %>%
#   select(-id, -First, -Last, -State)
# 
# set.seed(333)
# training1 <- createDataPartition(pred_data2$like_wvu, p = .75)$Resample1
# train_data1 <- pred_data2[training1,]
# test_data1 <- pred_data2[-training1, ]
# 
# set.seed(333)
# fit.rf1 <- train(All.Conf.NFL ~., data = train_data1, method="rf", trControl = fitControl1,
#                  family="binomial")
# 
# pr.rf1 <- predict(fit.rf1, test_data1)
# confusionMatrix(pr.rf1, test_data1$All.Conf.NFL)
# ```
# ```{r, echo = FALSE}
# varImp(fit.rf1)
```

##Graphically Analyze the "Like WVU" Category
I want to take the commits that went to the like_wvu category and gain insight into the makup of the 198 observations. In particular, I want to see if we can narrow down a geographic region or other valuable insight for the state of Georgia in terms of focusing our recruiting efforts. 

```{r, include = FALSE}
###Look at the like_wvu data to narrow down the seach
wv_data <- wv_rec %>%
  filter(year == 2016 | year == 2017 | year == 2018 | year == 2019 | year == 2020)

count_schools <- wv_data %>%
  count(High.School) %>%
  rename(num_commits = n)

wv_data <- left_join(wv_data, count_schools, by = "High.School") %>%
  filter(num_commits > 2 | Signed == "WVUN")
  

#get unique town names
geo_towns <- wv_data %>%
  filter(num_commits >1) %>%
  select(High.School) %>%
  distinct()

town <- c("Powder Springs", "Fairburn", "Fairburn", "College Park", "Ellenwood",
          "Suwanee", "Grayson", "Hoschton", "Stone Mountain", "Buford", "Alpharetta",
          "Richmond Hill")
towns2 <- c("Powder$20Springs", "Fairburn", "Fairburn", "College%20Park", "Ellenwood",
          "Suwanee", "Grayson", "Hoschton", "Stone%20Mountain", "Buford", "Alpharetta",
          "Richmond%20Hill")

geo_towns <- cbind(geo_towns, town)

#pull lat/long information for plotting towns
latlongs <- NULL
ll <- NULL
api_key <- "uCkPDPmacsZIPqhzzHdXsHTZkGSmrbTS"
for (i in towns2) {
  latlong_link <- paste0("https://api.tomtom.com/search/2/geocode/",i ,"%2CGA.json?storeResult=false&typeahead=false&limit=1&countrySet=US&key=",api_key)

  js <- fromJSON(latlong_link)
  lat <- js$results$position$lat
  lon <- js$results$position$lon
  
  ll <- cbind(i, lat,lon)
  latlongs <- rbind(latlongs, ll)
  Sys.sleep(1) #wait 1 seconds between requests so I don't time out the server
}    

all_latlong <- as.data.frame(latlongs) %>%
  mutate("town" = i) %>%
  select(-i)
all_latlong$town <- sub("%20", " ", all_latlong$town)
all_latlong$lat <- as.numeric(levels(all_latlong$lat))[all_latlong$lat]
all_latlong$lon <- as.numeric(levels(all_latlong$lon))[all_latlong$lon]


#put the lat longs in the dataset
geo_data <- left_join(geo_towns, all_latlong, by = "town") %>%
  unique()
geo_data_p <- as.data.frame(geo_data) #for plotting later
geo_data <- left_join(wv_data, geo_data, by = "High.School")
geo_data <- geo_data[complete.cases(geo_data),]

#get lat/long for Atlanta to plot on map
atl_link <- paste0("https://api.tomtom.com/search/2/geocode/Atlanta%2CGA.json?storeResult=false&typeahead=false&limit=1&countrySet=US&key=",api_key)

  js2 <- fromJSON(atl_link)
  lat2 <- js2$results$position$lat
  lon2 <- js2$results$position$lon

  atl_ll <- data.frame("city" = "Atlanta", "lat" = lat2, "lon" = lon2)
  atl_ll$city <- as.character(atl_ll$city)


```
```{r, include = FALSE}
#setup plot for Georgia
us <- map_data("state")
ga_map <- subset(us, region %in% "georgia")

#setup plot
ga_plot <- ggplot(ga_map, mapping = aes(x = long, y = lat), group = group) +
  coord_fixed(1.3) +
  geom_polygon(color = "black", fill = "gray") +
  geom_point(data = geo_data, aes(x=lon, y=lat, size = num_commits, color = Conf.)) +
  geom_text(x = lon2-.3, y= lat2+.2, label = "Atlanta") +
  labs(title = "Cities in GA with More than 1 Commits in 5 Years",
       subtitle="To Schools 'Like' WVU",
       fill="Total",
       plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust=0.5))
```

```{r, echo = FALSE, fig.align='center'}
ga_plot

```

```{r, include = FALSE}
#count number of commits to a college by town
qty_town_coll <- geo_data %>%
  group_by(town, Signed) %>%
  count(town) %>%
  rename(by_town = n)


#plot college location of commit by town
plot_college <- ggplot(qty_town_coll, aes(x=town, y=by_town, fill=Signed, label=by_town)) + 
  geom_bar(stat="identity") +
  geom_text(size=3, position = position_stack(vjust=0.5)) +
  labs(title = "Commits By Town Colored by College",
       subtitle = "Years 2015 - 2020",
       x="Town",
       y="Number of Commits",
       fill="College") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust=0.5),
        axis.text.x = element_text(angle=45, hjust = 1)) +
  scale_y_continuous(breaks = c(1:50))
```
```{r, echo = FALSE, fig.align='center'}
plot_college
```

```{r, include = FALSE}
#count number of commits positions by town
qty_pos_coll <- geo_data %>%
  group_by(town, Pos.) %>%
  count(Pos.) %>%
  rename(by_pos = n)
qty_pos_coll$pos  <- as.character(qty_pos_coll$Pos.)

plot_pos <- ggplot(qty_pos_coll, aes(x=town, y=by_pos, fill=pos, label=by_pos)) + 
  geom_bar(stat="identity") +
  geom_text(size=3, position = position_stack(vjust=0.5)) +
  labs(title = "Commits By Town Colored by Position",
       x="Town",
       y="Number of Commits",
       fill="Position") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust=0.5),
        axis.text.x = element_text(angle=45, hjust = 1)) +
  scale_y_continuous(breaks = c(1:50))
```
```{r, echo = FALSE, fig.align='center'}
plot_pos
```


Another plot showing signed by position
```{r, include = FALSE}
#count number of commits positions by town
qty_pos_signed <- geo_data %>%
  group_by(Signed, Pos.) %>%
  count(Pos.) %>%
  rename(by_pos = n)
qty_pos_signed$Pos.  <- as.character(qty_pos_signed$Pos.)

plot_pos_signed <- ggplot(qty_pos_signed, aes(x=Signed, y=by_pos, fill=Pos., label=by_pos)) + 
  geom_bar(stat="identity") +
  geom_text(size=3, position = position_stack(vjust=0.5)) +
  labs(title = "Commits Position By Signed Colored by Position",
       x="College",
       y="Number of Commits",
       fill="Position") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust=0.5),
        axis.text.x = element_text(angle=45, hjust = 1)) +
  scale_y_continuous(breaks = c(1:50))
```
```{r, echo = FALSE, fig.align='center'}
plot_pos_signed
```

##Plot Showing Trends over Time

```{r, include = FALSE}
by_time <- geo_data %>%
  group_by(year) %>%
  count(year) %>%
  rename(commits = n) %>%
  as.data.frame()

plot_by_time <- ggplot(by_time, aes(x=year, y=commits)) +
  geom_bar(stat = "identity") +
  labs(title = "Commits to 'Like WVU' by Year",
       x="Year",
       y="Number of Commits") +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust=0.5),
        axis.text.x = element_text(angle=45, hjust = 1))
```

```{r, echo = FALSE, fig.align='center'}
plot_by_time
```

##Where Did All-Conf / NFL Go?

```{r, include = FALSE}
#get nfl all-conf players
nfl <- recruit_all
nfl$All.Conf.NFL <- as.numeric(levels(nfl$All.Conf.NFL))[nfl$All.Conf.NFL]

#clean up data
nfl <- nfl %>%
  filter(All.Conf.NFL == 1) %>%
  select(-First, -Last, -State)

#count number of times a school produced nfl all-conf player
nfl_hs <- nfl %>%
  group_by(High.School) %>%
  count(High.School) %>% 
  rename(nfl_ac = n)

#join with 'like-wvu' data
top_hs <- left_join(geo_data, nfl_hs, by = "High.School")
top_hs <- top_hs[complete.cases(top_hs),]

#number of commits in the like-wvu data that went all-conf nfl
top_hs <- top_hs %>%
  select(High.School, nfl_ac, lat, lon) %>%
  distinct(High.School, .keep_all = TRUE)

#setup plot

plot_nfl <- ggplot(ga_map, mapping = aes(x = long, y = lat), group = group) +
  coord_fixed(1.3) +
  geom_polygon(color = "black", fill = "gray") +
  geom_point(data = top_hs, aes(x=lon, y=lat, size = nfl_ac, color = High.School)) +
  geom_text(x = lon2-.3, y= lat2+.2, label = "Atlanta") +
  labs(title = "Schools That Produced NFL Talent for Like-WVU",
       fill="High School",
       plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust=0.5))
```

##Plot Top High Schools by Players in NFL & All-Conf
I want to show which players went to "like-WVU" schools that went on to have great succcess. 
```{r, echo = FALSE, fig.align='center'}
plot_nfl

```

