---
title: "WVU Football - 2019 vTexas"
author: "Jeremy Harris"
date: "April 28, 2020"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

#WVU Football Analysis
I'm creating this analysis as a pipeline to view WVU Football information. I have started with just a single game here and the data is limited. I plan to expand this into a Shiny App so a user can select the year and game that they are interested in and have the results automatically appear.

In addition, I plan to compile multiple years for a head coach and try to gain further insight into tendencies related to the game current game situation. This is just a starting point and will be built upon further. 

##Texas @ WVU - 2019
The game I'm using to setup the model is week 6 of the 2019 season when Texas visited WVU. Once I get the analysis fully setup, this should work for any team that WVU played. 

##The Data
I'm using this website: https://collegefootballdata.com to gather data for all plays that took place during the game. They have an API located here: https://api.collegefootballdata.com/api/docs/?url=/api-docs.json#/plays/getPlays that I am using to pull the data directly into my file. 


```{r, include = FALSE}
#load libraries
library(dplyr)
library(jsonlite)
library(tidyverse)

#get data from API 
data_in <- fromJSON("https://api.collegefootballdata.com/plays?seasonType=regular&year=2019&week=6&team=West%20Virginia")

#set variables for team that WVU is playing -- Week 6 is vs TEXAS
opp <- "Texas"
opp_score <- "score_TEX"
```


##Clean Up Data
I take the data and drop fields that aren't needed, and rearrange the data into an order that makes sense - then I sort by period then minutes then seconds. Most of my code I dropped from the pdf printout but it available on my github page (http://github.com/jeremy-harris). I left this here to show some of what is taking place behind the scenes. 

```{r}
#convert score to be associated with team, not offense/defense
data_in <- data_in %>%
  mutate(score_WVU = ifelse(offense == "West Virginia", offense_score, defense_score), 
         opp_score = ifelse(offense == opp, offense_score, defense_score)) %>%
  mutate(clock.minutes = clock[[1]], clock.seconds = clock[[2]])

#group by period, then minute, then second
data_in <- data_in %>%
  select(-drive_id, -offense_score, -defense_score, -yards_to_goal, -offense_conference, -id,
         -clock, -defense_conference, -defense, -ppa) %>%
  select(score_WVU, opp_score, period, clock.minutes,
         clock.seconds, offense, yard_line, down, distance, yards_gained, play_type, play_text) %>%
  arrange(period, -clock.minutes)

```

##Visualize The Data
I create a few different plots here to show things that might be interesting. I don't have tons of granular detail like formations, blitzing, injuries, etc. - but I can show basic game stats which is what I try to do here. 

```{r, include = FALSE}
library(ggplot2)
library(gridExtra)

#Let's show play type by yards to go for each team
distance_data <- data_in %>%
  mutate(play = ifelse(str_detect(string=play_type, pattern="Rush"), "Rush",
                       ifelse(str_detect(string=play_type, pattern = "Pass*"), "Pass", "Other"))) %>%
  filter(play_type != "End of Half" & play_type != "End Period" & play_type != "Kickoff" &
           play_type != "Field Goal Good" & play_type != "Field Goal Missed" &
           play_type != "Penalty" & play_type != "Punt" & play_type != "Timeout" &
           play_type != "Sack")

gg_yards_wvu <- distance_data %>% filter(offense == "West Virginia") %>%
  ggplot(aes(distance)) +
  geom_density(aes(fill = factor(play)), alpha = 0.6) +
  labs(title = "Density Plot of Pass/Rush by Distance to Go for First Down",
       subtitle="West Virginia",
       x="Distance to Get First Down/Score",
       fill="Play Type") + 
  scale_fill_manual(values = c("#FFFF33", "#000072"))

gg_yards_opp <- distance_data %>% filter(offense == opp) %>%
  ggplot(aes(distance)) +
  geom_density(aes(fill = factor(play)), alpha = 0.6) +
  labs(title = "Density Plot of Pass/Rush by Distance to Go for First Down",
       subtitle=opp,
       x="Distance to Get First Down/Score",
       fill="Play Type")
```

```{r, echo = FALSE}
plot_yards_wvu <- gg_yards_wvu + theme(plot.margin = unit(c(1,1,0,2), "lines"))
plot_yards_opp <- gg_yards_opp + theme(plot.margin = unit(c(2,1,1,1), "lines"))
grid.arrange(plot_yards_wvu, plot_yards_opp)
```

This next plot shows the type of play by down per team. 

```{r, include = FALSE}
#Let's show play type down for each team
down_data <- data_in %>%
  mutate(play = ifelse(str_detect(string=play_type, pattern="Rush"), "Rush",
                       ifelse(str_detect(string=play_type, pattern = "Pass*"), "Pass", "Other"))) %>%
  filter(play_type != "End of Half" & play_type != "End Period" & play_type != "Kickoff" &
           play_type != "Field Goal Good" & play_type != "Field Goal Missed" &
           play_type != "Penalty" & play_type != "Punt" & play_type != "Timeout" &
           play_type != "Sack") %>%
  count(down, offense, play)

gg_down <- down_data %>%
  ggplot(aes(x = down, y = n, fill = play)) + 
  geom_bar(stat = "identity") + 
  facet_wrap( ~ offense) +
  labs(title = "Pass/Rush by Down per Team",
       subtitle=opp,
       x="Down",
       y = "Number of Plays",
       fill="Play Type") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = c("#FFFF33", "#000072"))
```
```{r, echo = FALSE}
gg_down  
```

So, clearly Texas favored the run and WVU favored the pass. Also, since punting and field goals have been removed from this graph - we can see that Texas didn't go for it on 4th down during this game. 

##Statistical Analysis
I want to take a look at the probability of a given play type using all variables. As this is a single game, there isn't much data to use in this instance. However, a longer term project can compile data for an entire season (or coaches career, etc.) to show tenancies given the data in the dataset. Again, this can be combined with things like formation type of offense and defense to get an even better insight. 

```{r, echo = FALSE}
library(caret)
#first, I need to create two variables to use for the model: "who's winning" & "by how much"
#I also create a "run = true" showing that the play is a run play, I'll predict on this
data_mod <- data_in %>%
   mutate(play = ifelse(str_detect(string=play_type, pattern="Rush"), "Rush",
                       ifelse(str_detect(string=play_type, pattern = "Pass*"), "Pass", "Other")), 
         wvu_up = ifelse(score_WVU>opp_score, "1", "0"), 
         amt_up = abs(score_WVU - opp_score)) %>%
  filter(play != "Other") %>%
  select(-play_type, -play_text)

#create factors for appropriate variables
cnames <- colnames(data_mod[c(3,4,5, 6, 8,11,12)])
data_mod[cnames] <- lapply(data_mod[cnames], factor)


#now I want to setup my dataset for modeling holding out only 10% of the data
data_wvu <- data_mod %>%
  filter(offense == "West Virginia") %>%
  select(-offense)


data_opp <- data_mod %>%
  filter(offense == opp) %>%
  select(-offense)

#setup cross-validation with 10 folds as there are 681 observations in my data
fitControl <- trainControl(method = "cv", number = 10)

#setup test and train variables with 200 variables in the test dataset
set.seed(333)
training1 <- createDataPartition(data_wvu$play, p = .9)$Resample1
train_data1 <- data_wvu[training1, ]
test_data1 <- data_wvu[-training1, ]

training2 <- createDataPartition(data_opp$play, p = .9)$Resample1
train_data2 <- data_opp[training2, ]
test_data2 <- data_opp[-training2, ]

#model data with random forest and tree classification
#Fit with Random Forest
set.seed(333)
fit.rf1 <- train(play ~., data = train_data1, method="rf", trControl = fitControl,
                 family="binomial")
set.seed(333)
fit.rf2 <- train(play ~., data = train_data2, method="rf", trControl = fitControl,
                 family="binomial")

#prediction models with results compared to produce the best model (hold out set of 10% + 10 fold cv)
#Predict with random forest
pr.rf1 <- predict(fit.rf1, test_data1)
pr.rf2 <- predict(fit.rf2, test_data2)
```

I've setup models (Random Forest & Class Trees) to predict both WVU and the opponent's play to be a Run or Pass. Let's compare our models and see how they perform.

```{r, echo = FALSE}
#Random Forest Confusion Matrix - WVU
cat("\nRandom Forest - WVU\n")
confusionMatrix(pr.rf1, test_data1$play)

#Random Forest Confusion Matrix - opp
cat("\nRandom Forest - opp\n")
confusionMatrix(pr.rf2, test_data2$play)
```


##What The Stats Mean
From my limited data, I was able to show a 66% accuracy predicting the play type when WVU had the ball and a 71% accuracy predicting the play type when the opponent had the ball. However, that's a little ambitious and pretty generic given that my test set of data only had 6 or 7 plays per team. 

For WVU I was able to predict 4 pass plays correctly but missed 2 rushing plays. In effect, my model predicted all plays to be passing which is a decent approach considering how much WVU passed in this game vs. rushing - but I wouldn't consider that necessarily a "prediction.". 

For Texas I was able to see a little more accuracy and my model correctly predicted the pass play as well as 4 rushing plays. This time the model favors rushing here because Texas ran the ball more than passing. This model predicted 2 rushing plays that were actually pass plays.

##What's Next?
I think that looking at more games with more plays will provide better results, or at least results that even out. I think I'd like to show more plots with more analysis automatically. 

Two things I really want to do is give the user access to change the team and game and automatically show the same results. The next thing I want to do is some sort of animation - maybe show the number scores animate over time...or, for a really tricky problem (that may not really be anything other than "cool") - animate the position of the ball for the entire game using the yard-line and yards gained per play. 

##Let Me Know!
If you have a stat that you want to see or something you'd like to know...please email me. I'll do my best to build it in the model. 

Thanks!

jeremy.scott.harris@gmail.com
